Numerical simulation and entropy dissipative cure of the carbuncle instability for the shallow water circular hydraulic jump

### 2020-11-23 numerics_paper/version2/numerical_chj.tex
Shorten material on approximate Riemann solvers.

\newcommand{\W}{{\mathcal W}}
\newcommand{\A}{{\mathcal A}}
\newcommand{\apdq}{\A^+ \!\!{\Delta} Q}
\newcommand{\amdq}{\A^- \!\!{\Delta} Q}
\newcommand{\imh}{{i-1/2}}
\newcommand{\iph}{{i+1/2}}
\newcommand{\bff}{{\bf f}}
\newcommand{\bfr}{{\bf r}}
\newcommand{\bfF}{{\bf F}}
\newcommand{\bfu}{{\bf u}}
\newcommand{\bfv}{{\bf v}}
\newcommand{\bfq}{{\bf q}}
\newcommand{\bfx}{{\bf x}}
\newcommand{\sgn}{{\rm sgn}}
\newcommand{\Rus}{{\rm Rus}}
\newcommand{\Roe}{{\rm Roe}}

@@

\section{Numerical methods}
In this work, we consider Riemann-based finite volume methods.
Considering a one dimensional hyperbolic conservation law,
the cell average of the numerical solution given by a
finite volume scheme (with forward Euler) can be written as follows:
\begin{align}\label{first-order_FV}
  \bfu_i^{n+1}=\bfu_i^n-\frac{\Delta t}{\Delta x}\left[\bfF_{i+1/2}-\bfF_{i-1/2}\right],
\end{align}
where $\bfF_{i+1/2}=\bfF_{i+1/2}(u^n)$ is a numerical flux; i.e., $\bfF_{i+1/2}$ is
a discretization of the flux $\bff(\bfu(x_{i+1/2},t^n))$.
This numerical flux is composed of two parts, a consistent term that approximates
the physical flux at the interface between cells $i$ and $i+1$, and a stabilization term.
For instance, the well-known Rusanov's numerical flux is given by
\begin{align}\label{rusanov_num_flux}
  \bfF_{i+1/2}(\bfu)=\frac{\bff(\bfu_{i+1})+\bff(\bfu_i)}{2} - \lambda_{i+1/2}(\bfu_{i+1}-\bfu_i),
\end{align}
where $\lambda_{i+1/2}$ is an upper bound on the wave speed associated with the Riemann problem
composed of the states $\bfu_i$ and $\bfu_{i+1}$.
%If $z=u$ in \eqref{rusanov_num_flux}
%; i.e., if the Rusanov's numerical flux is obtained considering
%the cell averages of the solution
%and if the time step is small enough,
Method \eqref{first-order_FV} with Rusanov's flux \eqref{rusanov_num_flux}
is positivity preserving and entropy stable; see for instance
\cite{perthame1996positivity,tadmor2003entropy,guermond2016invariant}
However, the accuracy of the numerical flux, and hence of the method, is
limited to first order.
Two alternatives are common to improve the accuracy properties of Rusanov's method.
The numerical flux $\bfF_{i+1/2}$ can be approximated based on high-order polynomial
reconstructions (or combinations of them) around the interface $i+1/2$.
For smooth solutions, this strategy leads to arbitrarily high-order
methods. Alternatively, second-order accuracy can be obtained by removing some of the
numerical dissipation introduced by the stabilization component of the numerical flux.
The second-order Lax-Wendroff method \cite{lax1959systems} is a popular example of this approach.
In this work we consider an adaptation of Lax-Wendroff method,
sometimes known as Lax-Wendroff-LeVeque (LWL) \cite{leveque1997wave,leveque2002finite},
that incorporates limiters into the second-order correction.
These high-order corrections to Rusanov's method require careful treatment to guarantee
the solution is still positivity preserving and/or entropy stable.

The accuracy of a given first- or second-order method depends heavily on the Riemann solver.
In particular, Rusanov's Riemann solver is highly dissipative. In contrast, the well-known
Roe's Riemann solver produces more accurate solutions.
%
As we discussed in the introduction, finite volume methods with
Roe's Riemann solvers are prone to the carbuncle instability for the CHJ.
In contrast, if we use Rusanov's
Riemann solver, no carbuncles appear; however, many other important and physical features of
the solution are dissipated, unless the grid is highly refined.
Our main objective is to study the CHJ and to propose a Riemann solver
that is carbuncle free but do not dissipate important features of the solution.
To do that we propose a Riemann solver that blends Rusanov's and Roe's solvers.
Our blending criterion is based on the residual of the entropy (in)equality,
which induces dissipation of entropy near discontinuities.
Blending these Riemann solvers has been proposed before
to solve the Euler equations, see e.g.
\cite{nishikawa2008very,wang2016developing,jaisankar2007diffusion,ohwada2018simple,deng2019new,ray2013entropy},
the shallow water equations, see e.g. \cite{bader2014carbuncle,kemm2014note},
and even the Navier-Stokes equations, see e.g. \cite{nishikawa2008very,ohwada2018simple}.
In these references, the authors use blending criteria different than ours.
Nevertheless, fixing the carbuncle instability by imposing entropy stability is not new.
For example, in \cite{ismail2009affordable,ismail2009proposed},
the authors developed carbuncle free solvers for the Euler equations
via dissipation of entropy.

\subsection{Lax-Wendroff-LeVeque method}\label{sec:lax-wendroff-leveque}
Let us now review the second-order correction to method \eqref{first-order_FV} implemented in
Clawpack \cite{clawpack} and PyClaw \cite{pyclaw-sisc}.
We refer to \cite{leveque1997wave, leveque2002finite} for a detailed description of the method.
We consider only a one-dimensional problem; in the
numerical experiments in \S\ref{sec:num} we use second-order Strang's splitting \cite{strang1968construction}
to extend the method to multiple dimensions.

The finite volume method \eqref{first-order_FV} can be written in terms of right-
and left-going waves emanating from the interfaces $i-1/2$ and $i+1/2$, respectively.
The net contribution of these waves is called right- and left-going fluctuations.
Upon defining $z^+:=\frac{1}{2}(z+|z|)$ and $z^-:=\frac{1}{2}(z-|z|)$,
the fluctuations are given by
\begin{align}\label{fluct}
  %\A^+\Delta Q_{i-1/2} := \frac{1}{2}\sum_p\left(s_{i-1/2}^p+|s_{i-1/2}^p|\right)\W_{i-1/2}^p, \quad
  %\A^-\Delta Q_{i+1/2} := \frac{1}{2}\sum_p\left(s_{i+1/2}^p-|s_{i+1/2}^p|\right)\W_{i+1/2}^p,
  \apdq_\imh := \sum_p\left(s_{i-1/2}^p\right)^+\W_{i-1/2}^p, \qquad
  \amdq_\iph := \sum_p\left(s_{i+1/2}^p\right)^-\W_{i+1/2}^p, \\
\end{align}
where $\W_{i\pm 1/2}^p$ and $s_{i\pm 1/2}^2$ are the waves and their corresponding speeds, respectively.
The waves and their speed of propagation (to compute the fluctuations) are tightly related to Riemann solvers.
In \S\ref{sec:riemann_solvers} we review the definition of the waves and speeds associated with
Rusanov's and Roe's Riemann solvers.
Based on fluctuations, method \eqref{first-order_FV} can be written as
\begin{align}\label{first-order_via_fluct}
  u_i^{n+1}=u_i^n-\frac{\Delta t}{\Delta x}\left[\apdq_\imh+\amdq_\iph\right].
\end{align}

The second-order LWL method removes some of the numerical dissipation in
\eqref{first-order_via_fluct} via a flux correction at each interface. The method is written as follows:
\begin{align}\label{second-order_via_fluct}
  u_i^{n+1} = u_i^n-\frac{\Delta t}{\Delta x}\left[\A^+\Delta Q_{i-1/2}+\A^-\Delta Q_{i+1/2}\right]
  -\frac{\Delta t}{\Delta x}\left[\tilde{F}_{i+1/2}-\tilde{F}_{i-1/2}\right],
\end{align}
where
\begin{align*}
  \tilde{F}_{i\pm 1/2}=\frac{1}{2}\sum_p|s_{i\pm 1/2}^p|\left(1-\frac{\Delta t}{\Delta x}|s_{i\pm 1/2}^p|\right)\tilde\W_{i\pm 1/2}^p
\end{align*}
are the flux corrections at the interfaces $i\pm 1/2$. Here $\tilde{\W}_{i\pm 1/2}^p$ is a
TVD-limited version of $\W_{i\pm 1/2}^p$.
In this work, we consider the min-mod limiters, we refer to \cite{leveque1997wave} for more details.
If no limiters are used
($\tilde \W_{i\pm 1/2}^p = \W_{i\pm 1/2}^p$), \eqref{second-order_via_fluct} becomes
the standard second-order Lax-Wendroff scheme, which in terms of fluctuation-like quantities
is given by
\begin{align*}
  u_i^{n+1}=u_i^n-\frac{\Delta t}{\Delta x}
\left[
  \frac{1}{2}\left(s_{i-1/2}^p+\frac{\Delta t}{\Delta x}|s_{i-1/2}^p|^2\right)\W_{i-1/2}^p
  +
  \frac{1}{2}\left(s_{i+1/2}^p-\frac{\Delta t}{\Delta x}|s_{i+1/2}^p|^2\right)\W_{i+1/2}^p
\right].
\end{align*}
To implement method \eqref{second-order_via_fluct},
we need to choose a Riemann solver. Although the scheme is second-order
regardless of the Riemann solver, the accuracy and robustness depends upon the
Riemann solver. The approximate Riemann solver based on Rusanov's flux is robust and known to be carbuncle free.
However, it adds large amounts of numerical diffusion that, for relatively coarse grids,
might dissipate other physical and important features of the solution. If the approximate Riemann solver
is based on Roe's average, the amount of numerical diffusion is highly reduced; however,
such an approach is prone to develop the carbuncle instability for certain problems.
Many carbuncle fixes to Roe's solver have been proposed for the Euler equations.
In the context of the shallow water equations, we review in \S\ref{sec:riemann_solvers}
the Riemann solver proposed in \cite{kemm2014note}.

\section{Riemann solvers}\label{sec:riemann_solvers}

For the shallow water equations in two-dimensions, the structure of the exact (and some
approximated) Riemann solvers contains three waves. This is the case, for example, of the well-known
Roe's Riemann solver, see for instance \cite{roe1981approximate} and
\cite[\S 15.3.3 and \S 21.7]{leveque2002finite}.
Simpler approaches are also possible.
For instance, methods \eqref{first-order_via_fluct} and \eqref{second-order_via_fluct}
with Rusanov's numerical flux are based on only two waves that
propagate with the same speed in opposite directions.
We refer to \cite{ketcheson2020riemann} and references therein for
a full and detailed review of these two Riemann solvers.

\subsection{Roe's average}\label{sec:roe}
The Roe's Riemann solver is based on evaluating the flux Jacobian using Roe's average
\begin{align}\label{roe_average}
  \bar h_{i-1/2}=\frac{1}{2}(h_{i-1}+h_i), \qquad
  \hat u_{i-1/2}=\frac{\sqrt{h_{i-1}}u_{i-1}+\sqrt{h_i}u_i}{\sqrt{h_{i-1}}+\sqrt{h_i}}, \qquad
  \hat v_{i-1/2}=\frac{\sqrt{h_{i-1}}v_{i-1}+\sqrt{h_i}v_i}{\sqrt{h_{i-1}}+\sqrt{h_i}}.
\end{align}
Let
\begin{align*}
\hat A_{i-1/2} =
\begin{bmatrix}
  0 & 1 & 0 \\
  -\hat u_{i-1/2}^2+g\bar h_{i-1/2} & 2\hat u_{i-1/2} & 0 \\
  -\hat u_{i-1/2} \hat v_{i-1/2} & \hat v_{i-1/2} & \hat u_{i-1/2}
\end{bmatrix}
\end{align*}
be the flux Jacobian evaluated using Roe's average at the interface $i-1/2$.
The average \eqref{roe_average} is designed to guarantee:
i)  consistency of the approximate flux Jacobian
($\hat A_{i-1/2}\rightarrow \bff^\prime(\bfu(x_{i-1/2}))$ as $\bfu_i, \bfu_{i-1}\rightarrow \bfu(x_{i-1/2})$);
ii) hyperbolicity ($\hat A_{i-1/2}$ is diagonalizable with real eigenvalues);
and iii) conservation ($\hat A_{i-1/2}(\bfu_i-\bfu_{i-1})=\bff(\bfu_i)-\bff(\bfu_{i-1})$).
Let
\begin{align*}
  \hat \bfr^1_{i-1/2} =
  \begin{bmatrix}
    1 \\
    \hat u_{i-1/2}-\sqrt{g\bar h_{i-1/2}}\\
    \hat v_{i-1/2}
  \end{bmatrix},
  \qquad
  \hat \bfr^2_{i-1/2} =
  \begin{bmatrix}
    0\\
    0\\
    1
  \end{bmatrix},
  \qquad
  \hat \bfr^3_{i-1/2} =
  \begin{bmatrix}
    1 \\
    \hat u_{i-1/2}+\sqrt{g\bar h_{i-1/2}}\\
    \hat v_{i-1/2}
  \end{bmatrix}
\end{align*}
be the eigenvectors of $\hat A_{i-1/2}$.
The waves for the two-dimensional Riemann solver based on Roe's average are
$\W_{i-1/2}^p=\alpha_{i-1/2}^p\hat r_{i-1/2}^p$, with $p=1,2,3$. Here $\alpha_{i-1/2}^p$ is given by
\begin{align}\label{system_for_alphas}
  \left[\hat \bfr^1_{i-1/2} ~\hat \bfr^2_{i-1/2} ~\hat \bfr^3_{i-1/2}\right]
  \begin{bmatrix}
    \alpha^1_{i-1/2} \\
    \alpha^2_{i-1/2} \\
    \alpha^3_{i-1/2}
  \end{bmatrix}
  =\Delta Q_{i-1/2}:=\bfu_i-\bfu_{i-1}.
\end{align}
%where $\alpha_{i-1/2}^p$ is the $p$-th entry from the solution of
%\end{align*}
The speed of the waves is given by the eigenvalues of $\hat A_{i-1/2}$
\begin{align*}
  \hat\lambda_{i-1/2}^1=\hat u_{i-1/2}-g\sqrt{g\bar h_{i-1/2}}, \qquad
  \hat\lambda_{i-1/2}^2=\hat u_{i-1/2}, \qquad
  \hat\lambda_{i-1/2}^3=\hat u_{i-1/2}+g\sqrt{g\bar h_{i-1/2}}. \\
\end{align*}

The waves and speeds at interface $i+1/2$ are computed similarly.
The fluctuations are given by \eqref{fluct} with $s_{i\pm 1/2}^p=\hat\lambda_{i\pm 1/2}^p$
Finally, the first- and second-order methods based on Roe's Riemann solver are given by
\eqref{first-order_via_fluct} and \eqref{second-order_via_fluct}, respectively.
We close this review by remarking that the conservation property
$\hat A_{i-1/2}(\bfu_i-\bfu_{i-1})=\bff(\bfu_i)-\bff(\bfu_{i-1})$ implies that
\begin{align}\label{cons_roe}
  \A^-\Delta Q_{i-1/2}+\A^+\Delta Q_{i-1/2}=\sum_p \hat\lambda_{i-1/2}^p\W_{i-1/2}^p=\bff(\bfu_{i})-\bff(\bfu_{i-1}),
\end{align}
a property we will use in the next section.

\subsection{Rusanov's Riemann solver}\label{sec:rusanov}
The Rusanov's Riemann solver considers only two waves that emanate from each interface,
say for instance interface $i-1/2$, and travel with the same speed $\lambda_{i-1/2}^{\max}\geq 0$
but in opposite directions.
The speed of propagation $\lambda_{i-1/2}^{\max}$ is an upper bound on the maximum wave speed
associated with the corresponding Riemann problem between the states $\bfu_{i-1}$ and $\bfu_{i}$.
In this work, we define $\lambda_{i-1/2}^{\max}$ via the guaranteed upper bound from \cite[Prop. 3.7]{azerad2017well}.
The waves are given by
\begin{align}\label{rusanov_waves}
  \W_{i-1/2}^{1,\Rus}=\bfu_{i-1/2}-\bfu_{i-1}, \qquad
  \W_{i-1/2}^{2,\Rus}=\bfu_{i}-\bfu_{i-1/2},
\end{align}
where $\bfu_{i-1/2}$ is a middle state in the solution of the Riemann problem. The middle state is obtained
by imposing
\begin{align}\label{rus_cons}
  -\lambda_{i-1/2}^{\max}\W_{i-1/2}^{1,\Rus}+\lambda_{i-1/2}^{\max}\W_{i-1/2}^{2,\Rus} = \bff(\bfu_{i})-\bff(\bfu_{i-1}),
\end{align}
which guarantees the solver is conservative.
The fluctuations are given by \eqref{fluct} with $s_{i-1/2}^1=-\lambda_{i-1/2}^{\max}$
and $s_{i-1/2}^2=\lambda_{i-1/2}^{\max}$, and the first- and second-order methods based
on Rusanov's solver are given by \eqref{first-order_via_fluct} and \eqref{second-order_via_fluct}, respectively.

>>>>

\newcommand{\W}{{\mathcal W}}
\newcommand{\A}{{\mathcal A}}
\newcommand{\apdq}{\A^+ \!\!{\Delta} Q}
\newcommand{\amdq}{\A^- \!\!{\Delta} Q}
\newcommand{\Ft}{\tilde{F}}
\newcommand{\imh}{{i-1/2}}
\newcommand{\iph}{{i+1/2}}
\newcommand{\bff}{{\bf f}}
\newcommand{\bfr}{{\bf r}}
\newcommand{\bfF}{{\bf F}}
\newcommand{\bfu}{{\bf u}}
\newcommand{\bfv}{{\bf v}}
\newcommand{\bfq}{{\bf q}}
\newcommand{\bfx}{{\bf x}}
\newcommand{\sgn}{{\rm sgn}}
\newcommand{\Rus}{{\rm Rus}}
\newcommand{\Roe}{{\rm Roe}}

@@

\section{Numerical methods}
In this work we study the behavior of certain shock-capturing finite volume
methods based on the use of Riemann solvers.  For simplicity, we discuss these
solvers in the context of a 1-dimensional mesh.  
In the numerical experiments in \S\ref{sec:num} we use second-order Strang
splitting \cite{strang1968construction} to extend the method to multiple dimensions.

The numerical solution $u_i$
represents the average value over cell $i$, which extends from $x_\imh$ to $x_\iph$.
At each time step and at each interface $x_\imh$, we approximately solve the
Riemann problem with initial states $(u^n_{i-1},u^n_i)$.  The approximate
solution consists of a set of traveling discontinuities or waves $\W^p_\imh$
and corresponding speeds $s^p_\imh$.
We use the wave propagation framework developed by LeVeque \cite{leveque1997wave, leveque2002finite}
and employed in the Clawpack software \cite{clawpack,pyclaw-sisc}, which implements the LWL scheme
\begin{align}\label{second-order_via_fluct}
  u_i^{n+1} = u_i^n-\frac{\Delta t}{\Delta x}\left[\apdq_\imh+\amdq_\iph\right]
  -\frac{\Delta t}{\Delta x}\left[\tilde{F}_{i+1/2}-\tilde{F}_{i-1/2}\right],
\end{align}
Upon defining $z^+:=\frac{1}{2}(z+|z|)$ and $z^-:=\frac{1}{2}(z-|z|)$,
the fluctuations are given by
\begin{align}\label{fluct}
  \apdq_\imh := \sum_p\left(s_{i-1/2}^p\right)^+\W_{i-1/2}^p, \qquad
  \amdq_\iph := \sum_p\left(s_{i+1/2}^p\right)^-\W_{i+1/2}^p, \\
\end{align}
and represent the effect (to first
order accuracy) of waves coming from the solution of the Riemann problem at
each of the neighboring interfaces.  Meanwhile, $\Ft_{i\pm 1/2}$ are correction
fluxes that make the method second-order accurate:
\begin{align*}
  \tilde{F}_{i\pm 1/2}=\frac{1}{2}\sum_p|s_{i\pm 1/2}^p|\left(1-\frac{\Delta t}{\Delta x}|s_{i\pm 1/2}^p|\right)\tilde\W_{i\pm 1/2}^p.
\end{align*}
Here $\tilde{\W}_{i\pm 1/2}^p$ is a TVD-limited version of $\W_{i\pm 1/2}^p$.
We will sometimes work with
the first-order method obtained by setting the correction fluxes to zero:
\begin{align}\label{first-order_via_fluct}
  u_i^{n+1} = u_i^n-\frac{\Delta t}{\Delta x}\left[\apdq_\imh+\amdq_\iph\right]
\end{align}
which, for the conservative Riemann solvers we use herein, can also be written
as a first-order flux-differencing scheme:
\begin{align}\label{first-order_FV}
  \bfu_i^{n+1}=\bfu_i^n-\frac{\Delta t}{\Delta x}\left[\bfF_{i+1/2}-\bfF_{i-1/2}\right],
\end{align}
where $\bfF_{i+1/2}$ is a numerical approximation of the flux at $x_\iph$.

Next we briefly review two commonly-used Riemann solvers: those of Rusanov and
Roe.  Neither of these solvers deals with the carbuncle instability in fully
satisfactory way.  Rusanov's solver suppresses the carbuncle but is known to
also suppress related real instabilities, while Roe's solver exhibits carbuncles.
Later we will combine these two solvers in order to better deal with shock instability.
We refer to \cite{ketcheson2020riemann} and references therein for
a full and detailed description of these two Riemann solvers in the context of
the shallow water equations.

\subsection{Rusanov's Riemann solver}\label{sec:rusanov}
The Rusanov solver approximates the Riemann solution by with a single wave traveling in each
direction.  Both waves are assumed to travel with speed $\lambda^{\max}_\imh$, which is an
upper bound on the wave speeds appearing in the true solution of the Riemann problem.
In this work, we take $\lambda_{i-1/2}^{\max}$ to be the upper bound
from \cite[Prop. 3.7]{azerad2017well}.
The waves are then given by
\begin{align}\label{rusanov_waves}
  \W_{i-1/2}^{1,\Rus}=\bfu_{i-1/2}-\bfu_{i-1}, \qquad
  \W_{i-1/2}^{2,\Rus}=\bfu_{i}-\bfu_{i-1/2},
\end{align}
where the intermediate state $\bfu_{i-1/2}$ is determined by imposing conservation:
\begin{align}\label{rus_cons}
  -\lambda_{i-1/2}^{\max}\W_{i-1/2}^{1,\Rus}+\lambda_{i-1/2}^{\max}\W_{i-1/2}^{2,\Rus} = \bff(\bfu_{i})-\bff(\bfu_{i-1}).
\end{align}
The fluctuations are given by \eqref{fluct} with $s_{i-1/2}^1=-\lambda_{i-1/2}^{\max}$
and $s_{i-1/2}^2=\lambda_{i-1/2}^{\max}$, and the first- and second-order methods based
on Rusanov's solver are given by \eqref{first-order_via_fluct} and \eqref{second-order_via_fluct}, respectively.

\subsection{Roe's Riemann solver} \label{sec:roe}
The Roe Riemann solver is based on evaluating the flux Jacobian using Roe's average
\begin{align}\label{roe_average}
  \bar h_{i-1/2}=\frac{1}{2}(h_{i-1}+h_i), \qquad
  \hat u_{i-1/2}=\frac{\sqrt{h_{i-1}}u_{i-1}+\sqrt{h_i}u_i}{\sqrt{h_{i-1}}+\sqrt{h_i}}, \qquad
  \hat v_{i-1/2}=\frac{\sqrt{h_{i-1}}v_{i-1}+\sqrt{h_i}v_i}{\sqrt{h_{i-1}}+\sqrt{h_i}}.
\end{align}
The waves and speeds are given by the eigenvectors and eigenvalues of the approximate
flux Jacobian obtained using these averages, resulting in
\begin{align*}
  \hat\lambda_{i-1/2}^1=\hat u_{i-1/2}-g\sqrt{g\bar h_{i-1/2}}, \qquad
  \hat\lambda_{i-1/2}^2=\hat u_{i-1/2}, \qquad
  \hat\lambda_{i-1/2}^3=\hat u_{i-1/2}+g\sqrt{g\bar h_{i-1/2}}. \\
\end{align*}
and $\W_{i-1/2}^p=\alpha_{i-1/2}^p\hat r_{i-1/2}^p$, with
\begin{align*}
  \hat \bfr^1_{i-1/2} =
  \begin{bmatrix}
    1 \\
    \hat u_{i-1/2}-\sqrt{g\bar h_{i-1/2}}\\
    \hat v_{i-1/2}
  \end{bmatrix},
  \qquad
  \hat \bfr^2_{i-1/2} =
  \begin{bmatrix}
    0\\
    0\\
    1
  \end{bmatrix},
  \qquad
  \hat \bfr^3_{i-1/2} =
  \begin{bmatrix}
    1 \\
    \hat u_{i-1/2}+\sqrt{g\bar h_{i-1/2}}\\
    \hat v_{i-1/2}
  \end{bmatrix}
\end{align*}
and the factors $\alpha^p_\imh$ obtained by solving
\begin{align}\label{system_for_alphas}
  \left[\hat \bfr^1_{i-1/2} ~\hat \bfr^2_{i-1/2} ~\hat \bfr^3_{i-1/2}\right]
  \begin{bmatrix}
    \alpha^1_{i-1/2} \\
    \alpha^2_{i-1/2} \\
    \alpha^3_{i-1/2}
  \end{bmatrix}
  =\Delta Q_{i-1/2}:=\bfu_i-\bfu_{i-1}.
\end{align}

\section{An entropy-based blending of Rusanov and Roe}

As we discussed in the introduction, finite volume methods with
Roe's Riemann solvers are prone to the carbuncle instability for the CHJ.
In contrast, if we use Rusanov's
Riemann solver, no carbuncles appear; however, many other important and physical features of
the solution are dissipated, unless the grid is highly refined.
Our main objective is to study the CHJ and to propose a Riemann solver
that is carbuncle-free but does not dissipate important features of the solution.
To do that we propose a Riemann solver that blends Rusanov's and Roe's solvers.
Our blending criterion is based on the residual of the entropy (in)equality,
which induces dissipation of entropy near discontinuities.
Blending these Riemann solvers has been proposed before
to solve the Euler equations, see e.g.
\cite{nishikawa2008very,wang2016developing,jaisankar2007diffusion,ohwada2018simple,deng2019new,ray2013entropy},
the shallow water equations, see e.g. \cite{bader2014carbuncle,kemm2014note},
and even the Navier-Stokes equations, see e.g. \cite{nishikawa2008very,ohwada2018simple}.
In these references, the authors use blending criteria different than ours.
Nevertheless, fixing the carbuncle instability by imposing entropy stability is not new.
For example, in \cite{ismail2009affordable,ismail2009proposed},
the authors developed carbuncle free solvers for the Euler equations
via dissipation of entropy.

%We close this section by remarking that the fluctuations associated with Rusanov's solver
%can be expressed in terms of waves and speeds from Roe's solver. To see this, first consider \eqref{rus_cons}
%and rewrite the right-going fluctuation as follows:
%\begin{align}\label{rus_rs_aux}
%  \A^{+,\Rus}\Delta Q_{i-1/2}:=\lambda_{i-1/2}^{\max}\W_{i-1/2}^{2,\max}
%  = \bff(\bfu_{i})-\bff(\bfu_{i-1}) + \lambda_{i-1/2}^{\max}\left(\W_{i-1/2}^{1,\Rus}+\W_{i-1/2}^{2,\Rus}\right)
%  -\lambda_{i-1/2}^{\max}\W_{i-1/2}^{2,\Rus}.
%  %\A^{+,\Rus}\Delta Q_{i-1/2}=\frac{1}{2}\sum_p\left(\hat \lambda_{i-1/2}^p + \lambda_{i-1/2}^{\max}\right)\W_{i-1/2}^p
%\end{align}
%From \eqref{system_for_alphas} and \eqref{rusanov_waves} we get
%$\W^{1,\Rus}_{i-1/2}+\W^{2,\Rus}_{i-1/2}=\bfu_i-\bfu_{i-1}=\sum_p\W_{i-1/2}^p$.
%Using \eqref{cons_roe}, \eqref{rus_rs_aux} becomes
%\begin{align*}
%  \lambda_{i-1/2}^{\max}\W_{i-1/2}^{2,\max}
%  = \sum_p \left(\hat \lambda_{i-1/2}^p + \lambda_{i-1/2}^{\max}\right)\W_{i-1/2}^p
%  -\lambda_{i-1/2}^{\max}\W_{i-1/2}^{2,\Rus},
%\end{align*}
%which implies that
%\begin{subequations}\label{fluct_rus}
%\begin{align}
%  \A^{+,\Rus}\Delta Q_{i-1/2}=\frac{1}{2}\sum_p\left(\hat \lambda_{i-1/2}^p + \lambda_{i-1/2}^{\max}\right)\W_{i-1/2}^p,
%\end{align}
%and similarly,
%\begin{align}
%  \A^{-,\Rus}\Delta Q_{i+1/2}=\frac{1}{2}\sum_p\left(\hat \lambda_{i+1/2}^p - \lambda_{i+1/2}^{\max}\right)\W_{i+1/2}^p.
%\end{align}
%\end{subequations}

%\subsubsection{HLLEMCC solver}
%The HLLEMCC approximate Riemann solver for the shallow water equations was proposed
%in \cite{kemm2014note} specifically to deal with carbuncles.  The idea is to adjust
%the amount of dissipation applied to the shear wave, adding dissipation in regions
%where unphysical carbuncles would appear.  The approximate solution in this
%case consists of four waves; the faster waves use the HLLE wave speeds and carry
%jumps only in the depth and normal momentum, while the slower waves carry jumps
%only in the transverse momentum.  By adjusting the value of $\phi$, this solver
%can behave like the HLLE solver (if $\phi=1$) or like the Roe solver (if $\phi=0$).
%
%The middle states $q^1_m, q^2_m$ are the same as those used for the HLLE solver.
%Requiring conservation of the transverse momentum gives
%\begin{align*}
%  q^3_m = \frac{f^3_l - f^3_r - (\hat{v}-\phi\hat{c})q^3_l + (\hat{v}+\phi\hat{c})q^3_r}{2\phi\hat{c}}.
%\end{align*}
%Here $\hat{v}, \hat{c}$ are the Roe-average normal velocity and sound speed, respectively, whereas the function $\phi$ is given by
%\begin{equation}\label{if}
%\phi(\theta)=\min\left\lbrace 1,\, \left(\varepsilon\,\theta\,\max\left\lbrace0,\,1-Fr_{u}^{\alpha} \right\rbrace \right)   \right\rbrace^{\beta}
%\end{equation}
%where $Fr_{u}$ is the Froude number perpendicular to the cell face and $\varepsilon$, $\alpha$, and $\beta$ are parameters that have to be chosen appropriately. For the shallow water equations, authors in \cite{kemm2014note} propose to use $\varepsilon=10^{-3}$ and $\alpha=\beta=\frac{1}{3}$. The value of $\theta$ is calculated as the 2-norm of the residual in the Rankine-Hugoniot condition for the shear wave.
%
%The HLLEMCC solver follows the ideas of the Harten-Hyman entropy fix by splitting the shear wave with speed $\hat{u}$, into two waves moving with speeds $\hat{u}-\phi\hat{c}$ and $\hat{u}+\phi\hat{c}$ (see figure \ref{HLLEMCC}). Its implementation, in terms of the fluctuations, is performed by using the values $\hat{u}^{\pm}$ calculated with an absolute-value function defined as
%
%\begin{align*}
%\varphi(\lambda)=\frac{\vert \lambda-\phi(\theta)\hat{c}\,\vert+\vert \lambda+\phi(\theta)\hat{c}\,\vert}{2}
%\end{align*}
%
%\begin{figure}
%    \center
%    \includegraphics[width=0.6\textwidth]{figures/hllemcc.pdf}
%    \caption{Structure of the HLLEMCC approximate Riemann solution, with four waves.}
%    \label{HLLEMCC}
%\end{figure}

\subsection{The entropy residual}
Let $S_i$ be the set of cells containing cell $i$ and those that share a face with it.
Also let $\bfv(\bfu)=\eta^\prime(\bfu)$ be the entropy variable.
Based on \cite{guermond2011entropy}, we consider
\begin{align}\label{ent_residual}
  \int_{S_i} \bfv(\bfu) \cdot \left[ \frac{\partial \bfu}{\partial t} + \nabla\cdot \bff(\bfu)\right]d\bfx
  =\int_{S_i} \left[\frac{\partial \eta(\bfu)}{\partial t} + \bfv(\bfu) \cdot \nabla\cdot \bff(\bfu)\right] d\bfx
\end{align}
as a measurement of the entropy production in $S_i$.
To avoid the need to compute the time derivative of the entropy, we follow
\cite{guermond2018second, guermond2018well} and use
$\int \frac{\partial\eta(\bfu)}{\partial t}d\bfx=-\int \nabla\cdot\bfq(\bfu) d\bfx$,
which holds in smooth regions. Then we define
\begin{align}\label{Ri}
  \theta_i := \frac{R_i}{D_i},
\end{align}
with
\begin{align*}
  R_i=
  \left|\int_{S_i} \left[\bfv(\bfu) \cdot \nabla\cdot \bff(\bfu) - \nabla\cdot\bfq(\bfu) \right] d\bfx\right|,
\end{align*}
and $D_i$ being an upper bound of $R_i$ that acts as normalization so that $0\leq \theta_i\leq 1$.
Note that $R_i\approx 0$ if $\bfu$ is smooth in $S_i$.
In our implementation, we use
\begin{align*}
  R_i\approx \left|\bfv(\bfu_i)\cdot \int_{S_i}\nabla\cdot \bff(\bfu)d\bfx
  -\int_{S_i}\nabla\cdot\bfq(\bfu) d\bfx\right|,
  \quad
  %D_i = ||\bfv(\bfu_i)||_{\ell^2}\left|\left|\int_{S_i}\nabla\cdot\bff(\bfu)d\bfx\right|\right|_{\ell^2}
  %+\left|\int_{S_i}\nabla\cdot\bfq(\bfu)d\bfx\right|.
  D_i = \sum_{k=1}^{d+1}|\bfv_k(\bfu_i)|\left|\int_{S_i}\left(\nabla\cdot\bff(\bfu)\right)_kd\bfx\right|
  +\left|\int_{S_i}\nabla\cdot\bfq(\bfu)d\bfx\right|,
\end{align*}
where $({\bf z})_k$ denotes the $k$-th component of ${\bf z}\in\mathbb{R}^{d+1}$ and $d$ is the number of physical
dimensions. In the next section, we use $\theta_i$ to blend Rusanov's and Roe's Riemann solvers.

\subsection{Blended Riemann solver}\label{sec:blended_rs}
In this section, we use \eqref{Ri}, which is a normalized measurement of the entropy residual \eqref{ent_residual},
to blend Rusanov's and Roe's Riemann solvers. To do this, we define the blended numerical flux:
\begin{align}\label{num_flux_ev}
  F_{i+1/2} = \frac{\bff(\bfu_{i+1})+\bff(\bfu_i)}{2}
  - \frac{1}{2} \left( \theta_{i+1/2}\lambda_{i+1/2}^{\max} I + (1-\theta_{i+1/2})|\hat A_{i+1/2}| \right)(\bfu_{i+1}-\bfu_{i}).
\end{align}
Here $I\in\mathbb{R}^{3\times 3}$ is the identity matrix,
$\theta_{i+1/2}=\max\{\theta_{i+1},\theta_i\}$, $\lambda_{i+1/2}^{\max}$ is the
upper bound on the wave speed used in Rusanov's Riemann solver (see \S \ref{sec:rusanov}),
and $\hat A_{i+1/2}$ is Roe's averaged flux Jacobian (see \S \ref{sec:roe}).
